Deep mutational scanning of hemagglutinin helps identify evolutionarily successful human H3N2 influenza viruses

### 2018-01-30 paper/manuscript.tex
addd some Trevor edits

\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein.
 Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping,bhatt2011genomic}.
This evolution is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}, producing a spindly phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
Several lines of evidence indicate that the trunk lineage outcompetes other strains at least in part because it has higher fitness~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
A key challenge in the study of H3N2 evolution is to forecast which strain will dominate the upcoming influenza season, as this information can guide selection of a vaccine strain~\cite{morris2017predictive}.
In order to make these forecasts, it is important to identify the features that enable the trunk strain to persist as other strains die out.

@@

\subsection*{The experimental measurements can help discriminate between successful and unsuccessful influenza virus lineages}
A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories~\cite{lassig2017predicting,morris2017predictive}.
For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}.
Evolutionary forecasts may target fixation and loss of alleles in the viral population, which means distinguishing between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).

@@

However, there are differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}).
In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps because immune pressure sometimes drives the fixation of functionally less favorable mutations at epitope sites.
The fact that only epitope sites exhibit an increase in sequence preference over time can also be rationalized in terms of prior knowledge about protein evolution.
As proteins evolve, there can be epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
It is also known that immune pressure can increase the amount of epistasis among substitutions by selecting for combinations of functionally deleterious immune-escape mutations and counterbalancing secondary mutations~\cite{gong2014epistatically}.
We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}).
Such a trend is not apparent for non-epitope sites, probably because they are under less selection to fix adaptive mutations.

>>>>

\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein.
 Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping,bhatt2011genomic}.
This evolution is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}, producing a spindly phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
Several lines of evidence indicate that the trunk has higher fitness than other lineages~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
A key challenge in the study of H3N2 evolution is to forecast which strain will dominate the upcoming influenza season, as this information can guide selection of a vaccine strain~\cite{morris2017predictive}.
In order to make these forecasts, it is important to identify the features that enable the trunk strain to persist as other strains die out.

@@

\subsection*{The experimental measurements can help discriminate between successful and unsuccessful influenza virus lineages}
A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories~\cite{lassig2017predicting,morris2017predictive}.
For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}.
Evolutionary forecasts should be able to distinguish successful and unsuccessful viral lineages, which includes discriminating between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).

@@

However, there are differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}).
In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps because immune pressure sometimes drives the fixation of functionally less favorable mutations at epitope sites.
The fact that only epitope sites exhibit an increase in sequence preference over time can also be rationalized in terms of prior knowledge about protein evolution.
As proteins evolve, there can be epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
It is also known that immune pressure can increase epistasis among substitutions by selecting for combinations of functionally deleterious immune-escape mutations and counterbalancing secondary mutations~\cite{gong2014epistatically}.
We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}).
Such a trend is not apparent for non-epitope sites, perhaps because they are under less selection to fix adaptive mutations.
